---
title: "MIMIC-III Admissions Table"
output:
  html_document:
    toc: yes
    number_sections: true  
  html_notebook:
    toc: yes
---

efg | 2018-12-30

NOTES:  

* RPostgreSQL works fine with dates before 1970, but seems to truncate datetime timestamps to dates, and does not compute time intervals correctly.

* Rpostgres does NOT work with dates before 1970 (*fix expected in Feb. 2019*), but computes time intervals below correctly.

See info about `admissions` table in [online schema](https://mit-lcp.github.io/mimic-schema-spy/tables/admissions.html).

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

## Packages

```{r, comment=NA}
library(DBI)
library(RPostgres)   # be cautious with Postgres dates before 1970
library(tidyverse)
library(dbplyr, warn.conflicts=FALSE)
library(kableExtra)
library(ggplot2)
library(writexl)
```

## Helper function

```{r}
Show <- function(data)
{
  data            %>%
  kable("html")   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```

## Open database

```{r}
MimicDB <- dbConnect(RPostgres::Postgres(), 
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     options  = "-c search_path=mimiciii")

plotCaptionLeft  <- "MIMIC-III v1.4"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

# List tables in database

This works with RPostgres but not RPostgreSQL.

```{r, comment=NA}
dbListTables(MimicDB)
```

# List of fields in admissions table

```{r, comment=NA}
dbListFields(MimicDB, "admissions")
```

# Admissions table record count

tbl - create a table from a data source

in_schema - refer to a table in a schema.  Here schema is "mimiciii" and table is "admissions".

```{r}
admissions <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
```

```{r}
admissions           %>%
  summarize(n = n()) %>%
  Show()
```

# Sample admissions records

```{r, comment=NA}
admissions   %>%
  head(5)    %>%
  Show()
```

# Fields

## row_id

```{r}
admissions                                            %>%
  summarize(n = n(),
            distinctRowID = n_distinct(row_id),
            minRowID      = min(row_id, na.rm=TRUE),
            maxRowID      = max(row_id, na.rm=TRUE))  %>%
  collect()                                           %>%
  Show()
```

## subject_id

Foreign key. Identifies the patient.

Parent is patients table.

```{r}
admissions                                                   %>%
  summarize(n = n(),
            distinctSubjectID = n_distinct(subject_id),
            minSubjectID      = min(subject_id, na.rm=TRUE),
            maxSubjectID      = max(subject_id, na.rm=TRUE)) %>%
  Show()
```

There are two groups of `subject _id` ranges.

```{r}
admissions              %>%
  select(subject_id)    %>%
  collect               %>%
  ggplot(aes(x=subject_id))                                 +
  geom_histogram(fill="cornsilk", color="grey80", bins=100) +
  scale_x_continuous(breaks = 0:10 * 10000,
                     labels = function(n){format(n, scientific = FALSE)},
                     minor_breaks = NULL)                   +
  geom_vline(xintercept = 36500, color = "skyblue")         +
  labs(title = "Different ranges of MIMIC-III subject_ids",
       caption=c(plotCaptionLeft, plotCaptionRight))        +
  theme(plot.caption = element_text(hjust=c(0.0,1.0)))
```

The reason for a break in the subject_ids from ~33000 to 40000 is unclear.

Also unclear is why the counts are higher for the lower subject_ids and lower for the higher subject_ids.

## hadm_id

Primary key:  Identifies the hospital stay

```{r}
admissions                                         %>%
  summarize(n = n(),
            distinctHadm = n_distinct(hadm_id),
            minHadmID = min(hadm_id, na.rm=TRUE),
            maxHadmID = max(hadm_id, na.rm=TRUE))  %>%
  collect()                                        %>%
  Show()
```

## admittime, dischtime, deathtime

* Time of admission to the hospital.

* Time of discharge from the hospital.

* Time of death.

sample values

Unclear why time differences are in minutes here.

```{r}
admissions                                                   %>%
  select(admittime, dischtime, deathtime)                    %>%
  collect()                                                  %>%
  mutate(StayDays = as.double(dischtime - admittime) /1440)  %>%
  head(10)  %>%
  Show()
```

How does visit time vary?

Length of Stay = Discharge Time - Admit Time

```{r, fig.width = 8}
admissions %>%
  select(admittime, dischtime, deathtime)                     %>%
  collect()                                                   %>%
  mutate(StayDays = as.double(dischtime - admittime) / 1440)  %>%
  filter(StayDays > 0)                                        %>%

  ggplot(aes(x = StayDays, y = ..density..))                       +
    geom_histogram(fill="cornsilk", color="grey80", binwidth=0.25) +
    geom_density()                                                 +
    scale_x_log10(breaks = c(0.25, 0.50, 1, 2, 4, 7, 14, 28),
                  minor_breaks = NULL)                             +
    labs(title="MIMIC-III Length of Stay",
         x = "Stay Time [Days]",
         caption=c(plotCaptionLeft, plotCaptionRight))             +
    theme(plot.caption = element_text(hjust=c(0.0,1.0)))
```


deathtime is mostly missing

```{r}
admissions                  %>%
  filter(is.na(deathtime))  %>%
  count()                   %>%
  collect()                 %>%
  Show()
```

What are the death years?

```{r}
deathStats <-
  admissions                                    %>%
  select(deathtime)                             %>%
  filter(!is.na(deathtime))                     %>%
  collect()                                     %>%
  mutate(deathYear = as.numeric(substr(deathtime, 1, 4)))   %>%
  select(deathYear)                             %>%
  group_by(deathYear)                           %>%
  count()
```

```{r, comment=NA}
c(nrow(deathStats), sum(deathStats$n))
```

```{r}
ggplot(data=deathStats, aes(x=deathYear, y=n))                  +
  geom_bar(stat="identity", fill="cornsilk", colour="grey80")   +
  labs(title="MIMIC-III Death Year",
         x = "Death Year (time shifted)",
         y = "Number of Deaths",
         caption=c(plotCaptionLeft, plotCaptionRight))             +
    theme(plot.caption = element_text(hjust=c(0.0,1.0)))

```

```{r, comment=NA}
c(min(deathStats$deathYear), max(deathStats$deathYear))
```

Are deaths spread somewhat uniformly by month?  Roughly, yes.

```{r}
deathInfo <-
  admissions                                    %>%
  select(deathtime)                             %>%
  filter(!is.na(deathtime))                     %>%
  collect()                                     %>%
  mutate(deathYear  = as.numeric(substr(deathtime, 1, 4)),
         deathMonth = as.numeric(substr(deathtime, 6, 7)))
```

```{r, fig.width=8}
smoothScatter(deathInfo$deathYear,
              deathInfo$deathMonth,
              xlab = "Death Year",
              ylab = "Death Month")
```

The expectation is the above scatterplot should be mostly random.

## admission_type

```{r}
admissions                                               %>%
  group_by(admission_type)                               %>%
  summarize(n = n())                                     %>%
  Show()
```

## admission_location

```{r}
admissions                                               %>%
  group_by(admission_location)                           %>%
  summarize(n = n())                                     %>%
  Show()
```

## discharge_location

```{r}
admissions                                               %>%
  group_by(discharge_location)                           %>%
  summarize(n = n())                                     %>%
  Show()
```

## insurance

```{r}
admissions                                               %>%
  group_by(insurance)                                    %>%
  summarize(n = n())                                     %>%
  Show()
```

## language

```{r}
admissions                                               %>%
  group_by(language)                                     %>%
  summarize(n = n())                                     %>%
  Show()                                                 %>%
  scroll_box(height = "500px")
```

## religion

```{r}
admissions                                               %>%
  group_by(religion)                                     %>%
  summarize(n = n())                                     %>%
  Show()
```

## marital_status

```{r}
admissions                                               %>%
  group_by(marital_status)                               %>%
  summarize(n = n())                                     %>%
  Show()
```

## ethnicity

```{r}
admissions                                               %>%
  group_by(ethnicity)                                    %>%
  summarize(n = n())                                     %>%
  Show()                                                 %>%
  scroll_box(height = "500px")
```

## edregtime, edouttime

Thought to be ...

* Emergency Department Registation Time

* Emergency Department Out Time

```{r, comment=NA}
admissions   %>%
  select(edregtime, edouttime)  %>%
  head(5)    %>%
  collect()  %>%
  Show()

```

```{r}
admissions                  %>%
 filter(is.na(edregtime))   %>%
  count()                   %>%
  collect()                 %>%
  Show()
```


```{r}
admissions                  %>%
 filter(is.na(edouttime))   %>%
  count()                   %>%
  collect()                 %>%
  Show()
```

```{r}
admissions                   %>%
  filter(!is.na(edouttime))  %>%
  count()                    %>%
  collect()                  %>%
  Show()
```

## diagnosis

```{r}
diagnosisCounts <-
  admissions                                             %>%
  group_by(diagnosis)                                    %>%
  summarize(n = n())                                     %>%
  collect()

nrow(diagnosisCounts)
```

top 20

```{r}
diagnosisCounts     %>%
  arrange(desc(n))  %>%
  head(20)          %>%
  Show()
```

Write to file for further study.

A number of diagnosis fields have multiple diagnoses separate by ";".

```{r}
write_xlsx(diagnosisCounts, "Admission-Diagnosis.xlsx")
```

## hospital_expire_flag

```{r}
admissions                                               %>%
  group_by(hospital_expire_flag)                         %>%
  summarize(n = n())                                     %>%
  Show()
```

## has_chartevents_data

Hospital admission has at least one observation in the CHARTEVENTS table.

```{r}
admissions                                               %>%
  group_by(has_chartevents_data)                         %>%
  summarize(n = n())                                     %>%
  Show()
```

# Close database

```{r, comment=NA}
dbDisconnect(MimicDB)
```

*****

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:",
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
