---
title: "MIMIC-III Diagnoses Tables"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

efg | 2019-01-06

# Fact and Dimension tables

The MIMIC-III schema shows a fact table,  [diagnoses_icd](https://mit-lcp.github.io/mimic-schema-spy/tables/diagnoses_icd.html), with a record for each patient-admission-diagnosis.  The details of the related ICD 9 diagnosis codes are in the dimension table,  [d_icd_diagnoses](https://mit-lcp.github.io/mimic-schema-spy/tables/d_icd_diagnoses.html).

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

```{r, comment=NA}
plotCaptionLeft  <- "MIMIC-III v1.4"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

## Packages 

```{r, comment=NA}
library(DBI)
library(RPostgres)
library(tidyverse)
library(dbplyr, warn.conflicts=FALSE)
library(kableExtra)
library(ggplot2)
library(writexl)
```

## Helper function

```{r}
Show <- function(data, caption=NULL)
{
  data                            %>%
  kable("html", caption=caption)  %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```

## Open database

```{r, comment=NA}
MimicDB <- dbConnect(RPostgres::Postgres(), 
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     bigint   = "integer",                      # RPostgres
                     options  = "-c search_path=mimiciii")  

dimDiagnoses   <- tbl(MimicDB, in_schema("mimiciii", "d_icd_diagnoses"))
factDiagnoses  <- tbl(MimicDB, in_schema("mimiciii", "diagnoses_icd"))

factAdmissions <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
factPatients   <- tbl(MimicDB, in_schema("mimiciii", "patients"))
```

# Tables

## Dimension table:  d_icd_diagnoses

### Fields

```{r, comment=NA}
dbListFields(MimicDB, "d_icd_diagnoses")
```

### Record count

```{r, comment=NA}
dimDiagnoses  %>%
  count()     %>%
  Show()
```

The schema says this table has 14,710 records, but the raw data file D_ICD_DIAGNOSES.csv has 14,567 records.  The value above is correct.

### Sample records

```{r, comment=NA}
dimDiagnoses          %>% 
  arrange(icd9_code)  %>%
  head(10)            %>%
  Show(caption = "Samle Records from Diagnoses Dimension Table, d_icd_diagnoses")
```

## Fact table:  diagnoses_icd

### Fields

```{r, comment=NA}
dbListFields(MimicDB, "diagnoses_icd")
```

### Record count

```{r, comment=NA}
factDiagnoses %>%
  summarize(n = n(), 
            diagnosesCodes = n_distinct(icd9_code)) %>%
  Show()
```

A number of fact records are missing icd9_code values

```{r, comment=NA}
factDiagnoses               %>%
  select(icd9_code)         %>%
  collect()                 %>%
  summarize(icd9NAs = sum(is.na(icd9_code)))  %>%
  Show()
```

### Sample records

```{r, comment=NA}
factDiagnoses  %>% 
  head(5)      %>%
  Show(caption = "Sample Records from Diagnoses Fact Table, diagnoses_icd")
```

### Fields

#### seq_num:  diagnosis priority

```{r, comment=NA}
priorityCounts <-
  factDiagnoses      %>%
  group_by(seq_num)  %>%
  count()            %>%  
  collect()

priorityCounts       %>%
  Show()             %>%
  scroll_box(height = "400px")
```

Sequence 1 is the primary code.

```{r, fig.wdith=6, fig.height=4}
ggplot(data = priorityCounts, aes(x = seq_num, y = n))  +
  geom_bar(stat = "identity", fill = "cornsilk")        +
  scale_y_continuous(labels = scales::comma)            +
  labs(title = "Diagnosis Priority",
       subtitle = paste(format(sum(priorityCounts$n, na.rm=TRUE), 
                               big.mark = ","), "fact records"),
       x = "Diagnosis Priority (seq_num)",
       y = "Counts",
       caption=c(plotCaptionLeft, plotCaptionRight))    +
  theme(plot.caption = element_text(hjust=c(0.0,1.0)))
  
```


## Join diagnoses tables

Add dimension information to fact table so word descriptions can be viewed instead of codes.

### inner_join

```{r, comment=NA}
factDiagnoses                                 %>% 
  inner_join(dimDiagnoses, by = "icd9_code")  %>%
  select(-starts_with("row_id."))             %>%
  head(5)                                     %>%
  collect()                                   %>%
  Show(caption = "Sample Diagnoses Records (Fact + Dimension)")
```

Why were so many fact records lost in join?

```{r, comment=NA}
factDiagnoses                                 %>% 
  inner_join(dimDiagnoses, by = "icd9_code")  %>%
  count()                                     %>%
  Show()
```

```{r}
dimICD9codes <- 
  dimDiagnoses       %>%
  select(icd9_code)  %>%
  distinct()         %>%
  arrange(icd9_code) %>%
  collect()

nrow(dimICD9codes)
```

```{r}
factICD9codes <- 
  factDiagnoses      %>%
  select(icd9_code)  %>%
  distinct()         %>%
  arrange(icd9_code) %>%
  collect()

nrow(factICD9codes)
```

Fact codes not in dimension list

The dimension table could likely be supplemented with information about ICD-9-CM codes from the [National Center for Health Statistics](https://www.cdc.gov/nchs/icd/icd9cm.htm).

```{r}
factNotInDim <- setdiff(factICD9codes$icd9_code, dimICD9codes$icd9_code)
factNotInDim
```

Many dimension records are not referenced by a fact table.

```{r}
dimNotInFact <- setdiff(dimICD9codes$icd9_code, factICD9codes$icd9_code)
length(dimNotInFact)
```

### left_join

Keep all diagnosis codes even if not in dimension table using a `left_join`.  This will create NULLs for diagnosis descriptions.

```{r, comment=NA}
factDiagnoses                                 %>% 
  left_join(dimDiagnoses, by = "icd9_code")   %>%
  count()                                     %>%
  Show()
```

So, normally we'll use a left_join with `dimDiagnoses` and missing labels will be NULL.

# Summaries

## Summarize Diagnoses Counts

```{r, comment=NA}
diagnosesCounts <- 
  factDiagnoses                     %>% 
  filter(!is.na(icd9_code))         %>%
  group_by(icd9_code)               %>%
  count()                           %>%
  ungroup()                         %>%                   
  left_join(dimDiagnoses, 
             by = "icd9_code")      %>%
  select(n, everything(), -row_id)  %>%
  arrange(desc(n))                  %>%
  collect()
 
nrow(diagnosesCounts)
```

Many of the ICD 9 diagnoses codes in the dimension table are never referenced.

### Top 10

```{r}
diagnosesCounts %>% head(10) %>% Show()
```

## Summarize Diagnoses as Primary or Secondary

```{r, comment=NA}
diagnosesPrimaryCounts <- 
  factDiagnoses                     %>% 
  filter(!is.na(icd9_code))         %>%
  mutate(priority = ifelse(seq_num == 1, 
                           "Primary", 
                           "Secondary"))  %>%
  group_by(icd9_code, priority)     %>%
  count()                           %>%
  ungroup()                         %>%
  collect()                         %>%
  spread(priority, n, fill=0)       %>%
  mutate(n = Primary + Secondary)   %>%
  left_join(dimDiagnoses, 
            by = "icd9_code",
            copy = TRUE)            %>%
  select(n, Primary, Secondary, 
         everything(), -row_id)     %>%
  arrange(desc(n))                 
 
nrow(diagnosesPrimaryCounts)
```

### Top 10 Primary Diagnoses

```{r, comment=NA}
diagnosesPrimaryCounts    %>%
  arrange(desc(Primary))  %>%
  head(10)                %>%
  Show()
```

Save for further study

```{r, comment=NA}
write_xlsx(diagnosesPrimaryCounts, "MIMIC-III-Diagnoses-Counts.xlsx")
```

## Summarize Diagnoses by Age Intervals

```{r, comment=NA}
admitAges <-
  factDiagnoses                               %>% 
  inner_join(factPatients,
             by = "subject_id")               %>%
  inner_join(factAdmissions,
             by = c("subject_id",
                    "hadm_id"))                           %>%
  select(subject_id, hadm_id, icd9_code, dob, admittime)  %>%
  collect()                                               %>%
  mutate(    # collect from database and use R from mutate
          delta         = admittime - dob,
          admitAgeYears = as.numeric((admittime - dob)) / ( 365.25 * 86400),  # seconds
          ageInterval   = pmin(90, 10 * admitAgeYears %/% 10)    # age decades; 90 is 90+
          )                                                  

nrow(admitAges)
```

Sample records

```{r}
admitAges %>% 
  head(2) %>% 
  Show()
```

Counts by age intervals

```{r, comment=NA}
table(admitAges$ageInterval)
```

Interval percentages

```{r, comment=NA}
round(100 * table(admitAges$ageInterval) / nrow(admitAges), 2)
```

Here "0" means 0-9, "10" means 10-19, ..., "80"" means 80-89, "90"" means 90+

Side note:  Decile intervals -- if balanced groups are wanted

```{r, comment=NA}
quantile(admitAges$admitAgeYears,
         c(0.00, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1.00))  %>% round(1)
```

Admit ages are skewed to 50+ intervals

### Admit Age Density

```{r}
ggplot(admitAges, aes(x = admitAgeYears, y = ..density..))             +
    geom_histogram(fill="cornsilk", color="grey80", binwidth=1) +
    geom_density()                                              +
    xlim(0, 100)  + 
    labs(title="MIMIC-III Admit Age",
         x = "Admit Age [Years]",                        
         caption=c(plotCaptionLeft, plotCaptionRight))          +
    theme(plot.caption = element_text(hjust=c(0.0,1.0)))
```

### Counts by diagnosis and age interval

```{r, comment=NA}
admitAgesCounts <-
  factDiagnoses                                 %>% 
  inner_join(factPatients,
             by = "subject_id")                 %>%
  inner_join(factAdmissions,
             by = c("subject_id",
                    "hadm_id"))                 %>%
  select(subject_id, hadm_id, icd9_code, dob, admittime)  %>%
  collect()                                               %>%   # extract data from DB
  mutate(                                                       # normal R processing 
          delta         = admittime - dob,
          admitAgeYears = as.numeric((admittime - dob)) / ( 365.25 * 86400),  # seconds
          ageInterval   = pmin(90, 10 * admitAgeYears %/% 10)    # age decades; 90 is 90+
        )                                       %>%  
  group_by(icd9_code, ageInterval)              %>%
  count()                                       %>%  # counts in long format
  ungroup()                                     %>%
  arrange(icd9_code, ageInterval)               %>%
  spread(ageInterval, n, fill=0)                %>%  # long to wide format
  mutate(RowTotal = rowSums(.[-1], na.rm=TRUE)) %>%
  select(icd9_code, RowTotal, everything())     %>%  # reorder variables
  arrange(desc(RowTotal))                       %>%  # descending order 
  left_join(dimDiagnoses, by = "icd9_code",          # add code labels
            copy = TRUE)                        %>%  # table to tibble
  select(-row_id)                               %>%  # remove a variable
  rename( "0s"= "0", "10s"="10", "20s"="20", "30s"="30",
         "40s"="40", "50s"="50", "60s"="60", "70s"="70",
         "80s"="80", "90+"="90")                     # slightly better names

nrow(admitAgesCounts)
```

```{r}
admitAgesCounts  %>%
  head(5)        %>%
  Show()
```


Add total row at bottom

```{r}
admitAgesCounts <-
  bind_rows(admitAgesCounts,
            bind_cols(icd9_code = "ColumnTotal",
                      admitAgesCounts %>%
                        summarize_if(is.numeric, sum, na.rm=TRUE)))
```


```{r}
admitAgesCounts  %>%
  tail(3)        %>%
  Show()
```

Save for further study

```{r, comment=NA}
write_xlsx(admitAgesCounts, "MIMIC-III-Diagnoses-Counts-by-Age-Intervals.xlsx")
```

**Close Database**

```{r, comment=NA}
dbDisconnect(MimicDB)
```

*****

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
