---
title: "MIMIC-III labevents table -- nearly 28 million records"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

efg | 2019-01-26

# labevents table

Online [info about the labevents table](https://mimic.physionet.org/mimictables/labevents/).  The [schema](https://mit-lcp.github.io/mimic-schema-spy/tables/labevents.html) shows the related tables.

From the online info:

*The LABEVENTS data contains information regarding laboratory based measurements. The process for acquiring a lab measurement is as follows: first, a member of the clinical staff acquires a fluid from a site in the patient’s body (e.g. blood from an arterial line, urine from a catheter, etc). Next, the fluid is bar coded to associate it with the patient and timestamped to record the time of the fluid acquisition. The lab analyses the data and returns a result within 4-12 hours.*

*Note that the time associated with this result is the time of the fluid acquisition, not the time that the values were made available to the clinical staff.*

*The labevents table contains both in-hospital laboratory measurements and out of hospital laboratory measurements from clinics which the patient has visited (since the patient is not “in” a hospital when visiting a clinic, these patients often referred to as “outpatients” and the data is often called “outpatient” data).*

**Laboratory measurements for outpatients do not have a HADM_ID.**

*Some items are duplicated between the labevents and chartevents tables. In cases where there is disagreement between measurements, labevents should be taken as the ground truth.*

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

```{r, comment=NA}
plotCaptionLeft  <- "MIMIC-III v1.4"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

## Packages 

```{r, comment=NA}
library(DBI)
library(RPostgres)
library(tidyverse)
library(dbplyr, warn.conflicts=FALSE)
library(kableExtra)
library(ggplot2)
library(writexl)
```

## Helper function

```{r}
Show <- function(data, caption=NULL)
{
  data                            %>%
  kable("html", caption=caption)  %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```

## Open database

```{r, comment=NA}
MimicDB <- dbConnect(RPostgres::Postgres(), 
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     bigint   = "integer",                      # RPostgres
                     options  = "-c search_path=mimiciii")  

factLabEvents <- tbl(MimicDB, in_schema("mimiciii", "labevents"))

factAdmissions  <- tbl(MimicDB, in_schema("mimiciii", "admissions"))
factPatients    <- tbl(MimicDB, in_schema("mimiciii", "patients"))

dimLabItems     <- tbl(MimicDB, in_schema("mimiciii", "d_labitems"))
```

# Sample records

## Lab Events Fact

```{r}
factLabEvents  %>%
  head()       %>%
  Show()
```

## Lab Items Dimension

```{r}
dimLabItems  %>%
  head()     %>%
  Show()
```

# Record Counts

## Lab Events Fact

```{r, comment=NA}
factLabEvents  %>%
  count()      %>%
  Show()
```

## Lab Items Dimension

```{r}
dimLabItems                                   %>%
  summarize(n  = n(),
            nItemid = n_distinct(itemid),
            nLOINC  = n_distinct(loinc_code),
            nLabel  = n_distinct(label))      %>%
  Show()
```




# Frequency Counts

Join tables:  factLabEvents joined with dimLabItems 

Some LOINCs have multiple units of measure, but sometimes units only vary by case.

## Lab Events by LOINC

itemid is unique.  loinc_code and label are not.

```{r, comment=NA}
labItemCounts <- 
  factLabEvents                                      %>%
  inner_join(dimLabItems, by = "itemid")             %>%
  
  group_by(loinc_code, itemid, label)                %>%
  summarize(n        = n(),
            nUnits   = n_distinct(valueuom),
            Units    = str_flatten(distinct(valueuom), 
                                 collapse="|"))      %>%
  ungroup()                                          %>%
  
  arrange(loinc_code, itemid)                        %>%
  select(n, everything())                            %>%  
  collect()
  
nrow(labItemCounts)
```

Most of the items in the dimension table were referenced.

List of items in dimension table not mentioned in fact table.

```{r}
notReferenced <-
  dimLabItems                              %>%
  anti_join(labItemCounts, by = "itemid",
            copy = TRUE)                   %>%
  collect()

nrow(notReferenced)
```

```{r}
notReferenced  %>%
  Show()
```

Write to file for further study.

```{r}
write_xlsx(labItemCounts, "Lab-Event-Item-Counts.xlsx")
```

## Top 25 lab events by item count

```{r}
labItemCounts        %>%
  arrange(desc(n))   %>%
  head(25)           %>%
  Show()
```

## Lab events with multiple units of measure

```{r}
multipleUnits <- 
  labItemCounts       %>%
  filter(nUnits > 1)  

nrow(multipleUnits)
```

```{r}
multipleUnits  %>%
  Show()
```

```{r}
write_xlsx(multipleUnits, "Lab-Event-Multiple-Units.xlsx")
```

# Unit conversion issues

```{r}
targetLab <- 
  factLabEvents                           %>%
  inner_join(dimLabItems, by = "itemid")  %>%
  select(-starts_with("row_id."))         %>%
  
  filter(itemid == 50804)                 %>%
  collect()
  
  
nrow(targetLab)
```

```{r}
targetLab  %>%
  head(3)  %>%
  Show()
```

```{r}
targetLab                                                                 %>%
  summarize(nNumber = sum( sapply(valuenum, FUN=is.numeric), na.rm=TRUE)) %>%
  Show()
```

```{r}
counts <- table(targetLab$valueuom, useNA = "ifany")
facetName <- paste0(names(counts), " [", counts, "]")
facetName
```

```{r}
variable_labeller <- function(variable, value)
{
  return(paste("Units = ", facetName[value]))
}
```

```{r, fig.width=8, fig.height=6}
ggplot(targetLab, aes(x = valuenum, y = ..density..))              +
    geom_histogram(fill="cornsilk", color="grey80", bins = 100)    +
    geom_density()                                                 +
    xlim(0, quantile(targetLab$valuenum, 1.00, na.rm=TRUE))        +
    labs(title=paste0("MIMIC-III Lab Event: ", targetLab$label),
         subtitle = paste0("Category: ", targetLab$category[1],
                           ", Fluid: ",  targetLab$fluid[1],
                           ", LOINC: ",  targetLab$loinc_code[1]),
         x = targetLab$label,
         caption=c(plotCaptionLeft, plotCaptionRight))             +
    theme(plot.caption = element_text(hjust=c(0.0,1.0)))          
```

```{r, fig.width=8, fig.height=8}
ggplot(targetLab, aes(x = valuenum, y = ..density..))              +
    geom_histogram(fill="cornsilk", color="grey80", bins = 100)    +
    geom_density()                                                 +
    xlim(0, quantile(targetLab$valuenum, 0.99, na.rm=TRUE))        +
    labs(title=paste0("MIMIC-III Lab Event: ", targetLab$label),
         subtitle = paste0("Category: ", targetLab$category[1],
                           ", Fluid: ",  targetLab$fluid[1],
                           ", LOINC: ",  targetLab$loinc_code[1]),
         x = targetLab$label,
         caption=c(plotCaptionLeft, plotCaptionRight))             +
    theme(plot.caption = element_text(hjust=c(0.0,1.0)))           +
    facet_wrap(~valueuom, 
               nrow=length(counts),
               labeller = as_labeller(variable_labeller))
```


**Close Database**

```{r, comment=NA}
dbDisconnect(MimicDB)
```

*****

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
