---
title: "MIMIC-III labevents table -- nearly 28 million records"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

efg | 2019-02-16


# Purpose

# Background

Swapna Abhyankar et al, [Standardizing clinical laboratory data for secondary use](https://www.sciencedirect.com/science/article/pii/S1532046412000652), *Jouornal of Biomedical Informatics*, 45(2012) 642-650.

Online [info about the labevents table](https://mimic.physionet.org/mimictables/labevents/).  The [schema](https://mit-lcp.github.io/mimic-schema-spy/tables/labevents.html) shows the related tables.

From the online info:

*The LABEVENTS data contains information regarding laboratory based measurements. The process for acquiring a lab measurement is as follows: first, a member of the clinical staff acquires a fluid from a site in the patient’s body (e.g. blood from an arterial line, urine from a catheter, etc). Next, the fluid is bar coded to associate it with the patient and timestamped to record the time of the fluid acquisition. The lab analyses the data and returns a result within 4-12 hours.*

*Note that the time associated with this result is the time of the fluid acquisition, not the time that the values were made available to the clinical staff.*

*The labevents table contains both in-hospital laboratory measurements and out of hospital laboratory measurements from clinics which the patient has visited (since the patient is not “in” a hospital when visiting a clinic, these patients often referred to as “outpatients” and the data is often called “outpatient” data).*

**Laboratory measurements for outpatients do not have a HADM_ID.**

*Some items are duplicated between the labevents and chartevents tables. In cases where there is disagreement between measurements, labevents should be taken as the ground truth.*

An earlier project, [MIMIC-III labevents table – First Look](labevents-FirstLook.htm) identified duplicate codes by case in several fields and recoding to resolve the problems.  The recoding below is based on that exploration.

[What LOINC is](https://loinc.org/get-started/what-loinc-is/).

# Exploration

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

```{r, comment=NA}
plotCaptionLeft  <- "MIMIC-III v1.4"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

### Packages 

```{r, comment=NA}
library(DBI)
library(RPostgres)
library(tidyverse)
library(dbplyr, warn.conflicts=FALSE)
library(kableExtra)
library(ggplot2)
library(writexl)
```

### Helper function

```{r}
Show <- function(data, caption="", bigMark="")
{
  data                                          %>%
  kable("html", 
        caption=caption,
        format.args = list(big.mark=bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```

### MIMIC-III database

```{r, comment=NA}
MimicDB <- dbConnect(RPostgres::Postgres(), 
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     bigint   = "integer",                      
                     options  = "-c search_path=mimiciii")
```

Database tables

```{r, comment=NA}

factLabEvents <- tbl(MimicDB, in_schema("mimiciii", "labevents"))
dimLabItems   <- tbl(MimicDB, in_schema("mimiciii", "d_labitems"))
```

### Recoding to fix case duplicates

With 32 GB of memory, let's bring tables into R memory tibbles for manipulation.

#### labevents

```{r, comment=NA}
fixedLabEvents <- 
  factLabEvents                                  %>%
  collect()                                      %>%
  
  mutate(
          valueuom = recode(valueuom,
                            "MEQ/L"   = "mEq/L",
                            "MG/DL"   = "mg/dL",
                            "MM HG"   = "mm Hg",
                            "MOSM/KG" = "mOsm/kg",
                            "ng/dl"   = "ng/dL",
                            "NG/DL"   = "ng/dL",
                            "ng/ml"   = "ng/mL",
                            "nG/mL"   = "ng/mL",
                            "nG/ML"   = "ng/mL",
                            "RATIO"   = "Ratio",
                            "SECONDS" = "sec",
                            "uG/DL"   = "ug/dL",
                            "ug/ml"   = "ug/mL",
                            "UNITS"   = "units",
                            .missing  = "<missing>"
                           )
        ) 

dim(fixedLabEvents)
```

#### d_labitems

```{r}
fixedLabItems <-
  dimLabItems  %>% 
  collect()    %>%  # Use R for the fixes                            
  
  mutate(
          category = recode(category,
                            "BLOOD GAS"  = "Blood Gas",
                            "CHEMISTRY"  = "Chemistry",
                            "HEMATOLOGY" = "Hematology"
                           ),
        
         fluid = recode(fluid,
                        "ASCITES"          = "Ascites",
                        "BLOOD"            = "Blood",
                        "CSF"              = "Cerebrospinal Fluid (CSF)",
                        "JOINT FLUID"      = "Joint Fluid",
                        "OTHER BODY FLUID" = "Other Body Fluid",
                        "STOOL"            = "Stool",
                        "URINE"            = "Urine"
                       ),
  
         # TdT is Terminal deoxynucleotidyl Transferase 
         label = recode(label,
                        "Mesothelial cells" = "Mesothelial Cells",
                        "TDT"               = "TdT" 
                       )
    
        )

nrow(fixedLabItems)
```

## Many lab events are missing units of measure

```{r}
fixedLabEvents                               %>%
  filter(valueuom == "<missing>")            %>%
  count()                                    %>%
  Show(bigMark = ",")
```

## LOINC missing or present?

A number of lab events are missing LOINC values.

```{r}
 fixedLabEvents                              %>%
  inner_join(fixedLabItems, by = "itemid")   %>%
  mutate(loincPresent = ifelse(is.na(loinc_code), 
                                     "LOINC missing",
                                     "LOINC present"))  %>%
  count(loincPresent)                        %>%
  Show(bigMark = ",")
```

For now, let's create additional summaries below, but be aware that one of the "distinct" LOINC values is likely a missing value.

## Category Summary

Join tables:  `factLabEvents` joined with `dimLabItems`

```{r, comment=NA}
categorySummary <-
  fixedLabEvents                              %>%
  inner_join(fixedLabItems, by = "itemid")    %>%

  group_by(category)                          %>%
  summarize(n      = n(),                     
            nFluid = n_distinct(fluid),
            nLOINC = n_distinct(loinc_code),
            nLabel = n_distinct(label))       %>%
  ungroup()  

categorySummary    %>%
  Show(bigMark = ",")
```

## Fluid Summary

```{r, comment=NA}
fluidSummary <-
  fixedLabEvents                                  %>%
  inner_join(fixedLabItems, by = "itemid")        %>%
  
  group_by(fluid)                                %>%
  summarize(n      = n(),                     
            nCategory = n_distinct(category),
            nLOINC    = n_distinct(loinc_code),
            nLabel    = n_distinct(label))       %>%
  ungroup()

fluidSummary   %>%
 Show(bigMark = ",")
```

## Fluid-Category Combinations

```{r, comment=NA}
fluidCategoryCombos <-
  fixedLabEvents                              %>%
  inner_join(fixedLabItems, by = "itemid")    %>%
  group_by(fluid, category)                   %>%
  count()                                     %>%
  ungroup()                                   %>%
  spread(category, n, fill = 0)               %>%
  arrange(fluid)                              %>%
  mutate(LabEvents = rowSums(.[-1], na.rm=TRUE))
```

Add Column Totals

```{r, comment=NA}
fluidCategoryCombos <- 
  bind_rows(fluidCategoryCombos,
            bind_cols(fluid = "TOTAL",
                      fluidCategoryCombos %>%
                        summarize_if(is.numeric, sum, na.rm = TRUE)))
```


```{r, comment=NA}

fluidCategoryCombos   %>%
  Show(bigMark = ",")
```

```{r}
write_xlsx(fluidCategoryCombos , "Lab-Event-Fluid-Category-Counts.xlsx")
```

View [Lab-Event-Fluid-Category-Counts.xlsx](Lab-Event-Fluid-Category-Counts.xlsx).

## Items and LOINCs

itemid is unique.  loinc_code and label are not.

Some LOINCs have multiple units of measure.

```{r, comment=NA}
labItemCounts <- 
  fixedLabEvents                                     %>%
  inner_join(fixedLabItems, by = "itemid")           %>%
  
  group_by(loinc_code, itemid, label)                %>%
  summarize(n        = n(),
            nUnits   = n_distinct(valueuom),
            Units    = str_flatten(unique(valueuom), 
                                 collapse="|"))      %>%
  ungroup()                                          %>%
  
  arrange(loinc_code, itemid)                        %>%
  select(n, everything())                            %>%  
  collect()
  
nrow(labItemCounts)
```

Most of the items in the dimension table were referenced.

List of items in dimension table not mentioned in fact table.

```{r}
notReferenced <-
  fixedLabItems                            %>%
  anti_join(labItemCounts, by = "itemid",
            copy = TRUE)                   %>%
  collect()

nrow(notReferenced)
```

```{r}
notReferenced                      %>%
  arrange(fluid, category, label)  %>%
  Show()
```

Write to file for further study.

```{r}
write_xlsx(labItemCounts, "Lab-Event-Item-Counts.xlsx")
```

View [Lab-Event-Item-Counts.xlsx](Lab-Event-Item-Counts.xlsx).

### Top 25 lab events by item count

```{r}
labItemCounts        %>%
  arrange(desc(n))   %>%
  head(25)           %>%
  Show()
```

### Lab events with multiple units of measure

```{r}
multipleUnits <- 
  labItemCounts       %>%
  filter(nUnits > 1)  

nrow(multipleUnits)
```

```{r}
multipleUnits  %>%
  Show()
```

```{r}
write_xlsx(multipleUnits, "Lab-Event-Multiple-Units.xlsx")
```

View [Lab-Event-Multiple-Units.xlsx](Lab-Event-Multiple-Units.xlsx).

### LOINCs with multiple itemids

```{r}
LOINCwithMultipleItems <-
  labItemCounts               %>% 
  filter(!is.na(loinc_code))  %>%
  group_by(loinc_code)        %>%
  summarize(n        = n(),
            nItemid  = n_distinct(itemid),
            itemids  = str_flatten(unique(itemid), 
                                 collapse=" | "))    %>%
  ungroup()                                          %>%
  filter(nItemid > 1)  

LOINCwithMultipleItems %>%
  Show()
```

```{r}
write_xlsx(LOINCwithMultipleItems, "Lab-Event-LOINC-Multiple-itemids.xlsx")
```

View [Lab-Event-LOINC-Multiple-itemids.xlsx](Lab-Event-LOINC-Multiple-itemids.xlsx).

**Close Database**

# Discussion

LOINC (Logical Observation Identifiers Names and Codes) is a way of standardizing clinical lab results.  The 2012 article by Swapna Abhyankar et al, "Standardizing clinical laboratory data for secondary use,"  explains the process developed using MIMIC-II, and carried forward to MIMIC-III:

*We were able to map 87% of the unique laboratory tests that cover 94% of the total number of laboratory tests results. Of the 13% of tests that we could not map, nearly 60% were due to test names whose real meaning could not be discerned and 29% represented tests that were not yet included in the LOINC table. These results suggest that LOINC codes cover most of laboratory tests used in critical care.*

# Conclusions

Over 20 `itemid`s have 500,000 records or more. Nearly 60 have 100,000 or more.

Categorizing all lab events by *category* and *fluid* is a useful breakdown.

"Blood" is the largest *fluid* grouping containing nearly 92% of all lab events roughly split between *Chemistry* and *Hematology* categoires.

*Urine* is the second largest *fluid* group but with about 7% of labe events.  The other 7 *fluid* groups are small.

A number of lab events have missing units or multiple units of measure for the same `itemid`.  Before using these records, one should probably verify if they have roughly the same distribution as the values with majority specified units.

Records with LOINCs with multiple `itemid`s should be explored to see if their distributions are roughly the same.

At this point the value of the lab events without LOINCs is not clear.

Exploration of the lab event value distributions might be interesting.

```{r, comment=NA}
dbDisconnect(MimicDB)
```

*****

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
