---
title: "MIMIC-III labevents table -- First Look"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0; 
}
</style>

efg | 2019-02-16

# Purpose

This project aims to identify problem data items in the `labevent` fact table as well as the `d_labitems` dimension table in the MIMIC-III database, and to create recoding that fixes them. This involves a mechanical look at the data fields without a serious attempt to understand the meaning of the data.

Future projects will summarize the fixed data.

# Background

Online [info about the labevents table](https://mimic.physionet.org/mimictables/labevents/).  

The [schema](https://mit-lcp.github.io/mimic-schema-spy/tables/labevents.html) shows the related tables.

# Exploration

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(cli.unicode = FALSE)
time.1 <- Sys.time()
```

```{r, comment=NA}
plotCaptionLeft  <- "MIMIC-III v1.4"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

### Packages 

```{r, comment=NA}
library(DBI)
library(RPostgres)
library(tidyverse)
library(dbplyr, warn.conflicts=FALSE)
library(kableExtra)
library(ggplot2)
library(writexl)
```

### Helper function

```{r}
Show <- function(data, caption="", bigMark="")
{
  data                                          %>%
  kable("html", 
        caption=caption,
        format.args = list(big.mark=bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```

### MIMIC-III database

```{r, comment=NA}
MimicDB <- dbConnect(RPostgres::Postgres(), 
                     dbname   = "mimic",
                     user     = Sys.getenv("MIMIC_User"),
                     password = Sys.getenv("MIMIC_Password"),
                     bigint   = "integer",                      # RPostgres
                     options  = "-c search_path=mimiciii") 
```

Database tables

```{r}
factLabEvents <- tbl(MimicDB, in_schema("mimiciii", "labevents"))
dimLabItems   <- tbl(MimicDB, in_schema("mimiciii", "d_labitems"))
```

## Lab Events Fact

### Sample Records

```{r}
factLabEvents  %>%
  head()       %>%
  Show()
```

### Record Counts

```{r, comment=NA}
factLabEvents                                    %>%
  mutate(upperUnits = str_to_upper(valueuom))    %>%
  summarize(
             n           = n(),
             nRow        = n_distinct(row_id),
             nSubject    = n_distinct(subject_id),
             nHadm       = n_distinct(hadm_id),
             nItem       = n_distinct(itemid),
             nCharttime  = n_distinct(charttime),
             nValue      = n_distinct(value),
             nValuenum   = n_distinct(valuenum),
             nValueuom   = n_distinct(valueuom),
             nUpperUnits = n_distinct(upperUnits),
             nFlag       = n_distinct(flag)
           )                                     %>%
  Show(bigMark = ",")
```

When `nUpperUnits` < `nValueuom`, some of the units differ by case.

Note the `n_distinct` counts performed on the database ignore nulls.  For example, there are really 3 `flag` states instead of the 2 shown above:

```{r}
factLabEvents     %>%
  group_by(flag)  %>%
  count           %>%
  ungroup()       %>%
  Show(bigMark = ",")
```

Below we will find that if the data are brought into R first, the distinct `flag` count is 3 as expected.

### Unit differences by case

```{r, comment=NA}
lookForMultipleVersions <-
  factLabEvents                                  %>%
  collect()                                      %>% # use R, not DB for str_flatten below
  mutate(upperUnits = str_to_upper(valueuom))    %>%
  group_by(upperUnits)                           %>%
  summarize(n             = n(),
            nValueuom     = n_distinct(valueuom),
            unitVariation = str_flatten(unique(valueuom), 
                                        collapse = " | "))  %>%
  ungroup()                                      
 
nrow(lookForMultipleVersions)
```

Identify variation in units of measure for given upperUnits.

```{r}
problemUnits <-
  lookForMultipleVersions  %>%
  filter(nValueuom > 1)

problemUnits  %>%
  Show()
```

```{r}
write_xlsx(problemUnits, "Problem-Units.xlsx")
```

### Use majority vote to pick preferred unit label

Look at counts for variations in units and let's pick "majority vote"" for preferred case

```{r, comment=NA}
voteScores <-
  factLabEvents                                        %>%
  mutate(upperUnits = str_to_upper(valueuom))          %>%
  filter(upperUnits %in% lookForMultipleVersions$upperUnits)  %>%
  group_by(upperUnits, valueuom)                       %>%
  count()                                              %>%
  ungroup()                                            %>%
  collect()

nrow(voteScores)
```

Show the "votes" for the problem cases

```{r}
problemUnits                                 %>%
  inner_join(voteScores, by = "upperUnits")  %>%
  rename(nUpper = n.x, n = n.y)              %>%
  Show()
```

[How to select the rows with maximum values in each group with dplyr?](https://stackoverflow.com/questions/24237399/how-to-select-the-rows-with-maximum-values-in-each-group-with-dplyr)

Probably should worry about ties here?  Do that later.

```{r, comment=NA}
preferredUnits <-
  voteScores            %>%
  group_by(upperUnits)  %>%
  slice(which.max(n))   %>%
  rename(PreferredUnit = valueuom)

nrow(preferredUnits)
```

### Setup translate table

```{r, comment=NA}
translateUnits <-
  voteScores       %>%
  select(-n)       %>%
  inner_join(preferredUnits, 
             by = "upperUnits")   %>%
  select(-n)

nrow(translateUnits)
```

```{r}
write_xlsx(translateUnits, "Translate-Units.xlsx")
```

Results for problem units

```{r}
problemUnits         %>%
  inner_join(translateUnits, by = "upperUnits") %>%
  select(-valueuom)  %>%
  distinct()         %>%
  Show()
```

Verify translate table removes unit case differences.

Use `left_join` here to preserve NULLs in units.

```{r, comment=NA}
factLabEvents                                    %>%
  left_join(translateUnits, 
            by = "valueuom",
            copy = TRUE)                         %>%
  rename(Units = PreferredUnit)                  %>%
  summarize(
             n           = n(),
             nRow        = n_distinct(row_id),
             nSubject    = n_distinct(subject_id),
             nHadm       = n_distinct(hadm_id),
             nItem       = n_distinct(itemid),
             nCharttime  = n_distinct(charttime),
             nValue      = n_distinct(value),
             nValuenum   = n_distinct(valuenum),
             nValueuom   = n_distinct(valueuom),
             nUnits      = n_distinct(Units),
             nFlag       = n_distinct(flag)
           )                                     %>%
  Show(bigMark = ",")
```

Here `nUnits` should match the original `nUpperUnits` value.

### Recoding to fix problems

Use dplyr's `mutate` to fix unit problems instead of a translate table.

```{r, comment=NA}
fixedLabEvents <- 
  factLabEvents                                  %>%
  collect()                                      %>%
  
  mutate(
          valueuom = recode(valueuom,
                            "MEQ/L"   = "mEq/L",
                            "MG/DL"   = "mg/dL",
                            "MM HG"   = "mm Hg",
                            "MOSM/KG" = "mOsm/kg",
                            "ng/dl"   = "ng/dL",
                            "NG/DL"   = "ng/dL",
                            "ng/ml"   = "ng/mL",
                            "nG/mL"   = "ng/mL",
                            "nG/ML"   = "ng/mL",
                            "RATIO"   = "Ratio",
                            "uG/DL"   = "ug/dL",
                            "ug/ml"   = "ug/mL",
                            "UNITS"   = "units"
                           )
        )                                        %>%
  
  mutate(upperUnits = str_to_upper(valueuom)) 
```

Verify fix

```{r}
fixedLabEvents                                   %>%
  summarize(
             n           = n(),
             nRow        = n_distinct(row_id),
             nSubject    = n_distinct(subject_id),
             nHadm       = n_distinct(hadm_id),
             nItem       = n_distinct(itemid),
             nCharttime  = n_distinct(charttime),
             nValue      = n_distinct(value),
             nValuenum   = n_distinct(valuenum),
             nValueuom   = n_distinct(valueuom),
             nUpperUnits = n_distinct(upperUnits),
             nFlag       = n_distinct(flag)
           )                                     %>%      
  Show(bigMark = ",")
```

Since `nValueuom` = `nUpperUnits`, there are no longer any unit differences based on case.

The distinct counts for `nHadm`, `nValue`, `nValuenum` and `nFlag` are all one greater here than the initial count shown above, since `n_distinct` on an R table includes NULL (NA) as one of the distinct values.

```{r}
fixedLabEvents    %>%
  group_by(flag)  %>%
  count           %>%
  ungroup()       %>%
  Show(bigMark = ",")
```

## Lab Items Dimension

### Sample Records

Sample records

```{r}
dimLabItems  %>%
  head()     %>%
  Show()
```

### Record Counts

```{r}
dimLabItems                                   %>%
  summarize(n  = n(),
            nItemid   = n_distinct(itemid),
            nLOINC    = n_distinct(loinc_code),
            nCategory = n_distinct(category),
            nFluid    = n_distinct(fluid),
            nLabel    = n_distinct(label))      %>%
  Show()
```

### category problems

```{r}
dimLabItems           %>%
  group_by(category)  %>%
  count()             %>%
  ungroup()           %>%
  arrange(category)   %>%
  Show()
```

### fluid problems

```{r}
dimLabItems        %>%
  group_by(fluid)  %>%
  count()          %>%
  ungroup()        %>%
  arrange(fluid)   %>%
  Show()
```

### label problems

```{r}
dimLabItems                                 %>%
  collect()                                 %>% # let R do heavy lifting
  mutate(upperLabel = str_to_upper(label))  %>% 
  
  group_by(upperLabel)                      %>%
  summarize(n      = n(),
            nLabel = n_distinct(label),
            labels = str_flatten(unique(label), 
                                 collapse = " | "))   %>%
  ungroup()                                 %>%
  
  filter(nLabel > 1)                        %>%          
  Show()
```

### Duplicate LOINCs?

Not all labs are assigned LOINCs

```{r}
duplicateLOINCs <-
  dimLabItems           %>%
  group_by(loinc_code)  %>%
  count()               %>%
  ungroup()             %>%
  filter(n > 1)         %>%
  arrange(loinc_code)

duplicateLOINCs  %>%
  Show()
```

```{r}
dimLabItems                                       %>%
  inner_join(duplicateLOINCs, by = "loinc_code")  %>%
  select(-n)                                      %>%
  Show()
```

Explore later if lab items with same LOINCs can be safely combined, and what can be done with lab items missing LOINCs.

### Recoding to fix problems

```{r}
fixedLabItems <-
  dimLabItems  %>% 
  collect()    %>%  # Use R for the fixes                            
  
  mutate(
          category = recode(category,
                            "BLOOD GAS"  = "Blood Gas",
                            "CHEMISTRY"  = "Chemistry",
                            "HEMATOLOGY" = "Hematology"
                           )
        )                                      %>%
  
  mutate(
         fluid = recode(fluid,
                        "ASCITES"          = "Ascites",
                        "BLOOD"            = "Blood",
                        "CSF"              = "Cerebrospinal Fluid (CSF)",
                        "JOINT FLUID"      = "Joint Fluid",
                        "OTHER BODY FLUID" = "Other Body Fluid",
                        "STOOL"            = "Stool",
                        "URINE"            = "Urine"
                       )
        )                                      %>%
  
  # TdT is Terminal deoxynucleotidyl Transferase 
  mutate(
         label = recode(label,
                        "Mesothelial cells" = "Mesothelial Cells",
                        "TDT"               = "TdT" 
                       )
    
        )
  
## Verification

fixedLabItems                                  %>%  
  summarize(n  = n(),
            nItemid   = n_distinct(itemid),
            nLOINC    = n_distinct(loinc_code),
            nCategory = n_distinct(category),
            nFluid    = n_distinct(fluid),
            nLabel    = n_distinct(label))    %>%   
  Show()
```

Verify case problems are resolved.

```{r}
fixedLabItems     %>%
  count(category) %>%
  Show()
```

```{r}
fixedLabItems  %>%
  count(fluid) %>%
  Show()
```

Verify no duplicate `label`s by case.

```{r}
fixedLabItems                               %>%
  mutate(upperLabel = str_to_upper(label))  %>% 
  
  group_by(upperLabel)                      %>%
  summarize(n      = n(),
            nLabel = n_distinct(label),
            labels = str_flatten(unique(label), 
                                 collapse = " | "))   %>%
  ungroup()                                 %>%
  
  filter(nLabel > 1)                        %>%          
  Show()
```

**Close Database**

```{r, comment=NA}
dbDisconnect(MimicDB)
```

# Discussion

Data cleanup is often easier in R than using database approaches, but memory limitations for R may force more processing on the database backend.

`n_distinct` counts run on the database ignore NULLs, but when run on an in-memory tibble R includes NULL (NA) as a distinct value.


# Conclusions

"Majority" vote was useful to decide which case variant should be used as the preferred unit of measure.

Several data fields need recoding as described above to use `labevents` in data summaries:

* `valueuom` (value unit of measure) in the `labevents` has 13 values that should be recoded for consistency.

*  Several values in the `label`,`category` and `fluid` fields in the `d_labitems` table should be recoded for consistency.

Additional exploration of `itemid`s with the same LOINC is needed to verify if they can be combined as a single source of data.

Additional exploration of 'itemid's without LOINC codes is needed to understand the utility of the data.

*****

```{r, comment=NA, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", 
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
